wordmaker

### 2020-07-20 wordmaker.Rmd
pre results edits

---
title: "An interdisciplinary examination of stress and injury occurrence in athletes"
runtitle: "Short Title"
documentclass: frontiersHLTH  # or frontiersHLTH, or frontiersFPHY
author:
  - name: Dr Harry Fisher
    affiliation: '1'
    etal: Fisher # First author's last name. 
    email: harryfisher21@gmail.com # Indicates corresponding Author
    institution: Cardiff Metropolitan University
    street: Cyncoed Road
    zip: CF23 6XD
    country: United Kingdom
  - name: Dr Marianne Gittoes
    affiliation: '1'
  - name: Professor Lynne Evans
    affiliation: '1'
  - name: Miss Leah Bitchell
    affiliation: '1'  
  - name: Dr Richard Mullen  
    affiliation: '2'
  - name: Dr Marco Scutari
    affiliation: '3'    
affiliation:
  - id: '1'
    department: School of Sport and Health Science
    institution: Cardiff Metropolitan University
    city: Cardiff
    country: United Kingdom
  - id: '2' 
    department: Division of Sport, Health & Exercise Sciences
    institution: Brunel University
    city: London
    country: United Kingdom
  - id: '3' 
    institution: Istituto Dalle Molle di Studi sull'Intelligenza Artificiale (IDSIA) 
    city: Manno
    country: Switzerland

output: rticles::frontiers_article
csl: frontiers.csl
bibliography: library.bib
---

@@

```{r include=FALSE}
options(scipen=999, digits = 2)
knitr::opts_chunk$set(echo = FALSE)
source("r-files/functions.R")
source("r-files/methods_code.R")
source("r-files/results_code.R")
```

# Introduction {-}

Over the last four decades sport related injuries have received increased research attention [@Ivarsson2017].
This attention is unsurprising given the high incidence [@Rosa2014; @Sheu2016],
and undesirable physical and psychological effects of sports injuries [@Leddy1994; @Brewer2012].
To mitigate against both the increasing incidence and	undesirable consequences of injury, research  has identified several psychological [@Slimani2018],
anatomical [@Murphy2003],
biomechanical [@Neely1998; @Hughes2014]
and environmental [@Meeuwisse2007]
factors associated with sports injury occurrence.
Indeed, several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence [@Kumar2001; @Meeuwisse2007; @Wiese-Bjornstal2009],
of which one of the most widely cited was developed by Williams and Anderson [Fig \ref{fig:fig1}; @Andersen1988; @Williams1998].

\begin{figure}[!h]
\caption{{\bf Stress and injury model (Williams and Andersen, 1998).}}
\label{fig:fig1}
\end{figure}

Williams and Andersen's [@Williams1998] model proposed that when faced with a potentially stressful athletic situation, an athlete’s personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) will contribute to their response, either interactively or in isolation.
Central to the model is the stress response, which reflects the bi-directional relationship between athletes’ appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of many stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater risk of injury.

Within Williams and Andersen’s [@Williams1998] model, major life events, a component of an athlete’s history of stressors, most consistently predicts injury occurrence [@Williams2007];
specifically, major life events with a negative, as opposed to positive, valence [@Passer1983a; @Maddison2005].
However, personality traits and coping resources have also been	found to predict injury, with for example, athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to athletes with the opposite profile. [@Smith1990; @Lavallee1996; @Ivarsson2010].
However, the amount of variance explained by the psychosocial factors proposed by the model has been modest, typically between 5 - 30% [@Galambos2005; @Ivarsson2010]; suggesting other factors are also likely to contribute to injury occurrence.

While the psychosocial factors proposed in Williams and Andersen’s [@Williams1998] model have received the most research attention, the mechanisms through which these factors are proposed to exert their effect have remained under-investigated in the literature. 
To elaborate, the  model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing.
However, to date, the research has largely focused on attentional deficits [@Andersen1999; @Rogers2005; @Wilkerson2012a; @Swanik2007].
For example, Andersen and Williams [@Andersen1999]
measured peripheral and central vision during high and low stress conditions and found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following season.
Indeed, Rodgers and Landers [@Rogers2005] supported Andersen and Williams's [@Andersen1999] earlier findings reporting that peripheral narrowing under stress mediated 8.1% of the relationship between negative life events and injury.
However, the remaining variance between negative life events and athletic injury through the other proposed mechanisms, such as increased muscle tension and reduced motor control, remains to be explored [cf. @Williams1998].

One challenge faced by researchers addressing the sports injury problem with a psychological lens is the multifactorial nature of injury, and the possible contribution of other non-psychological factors to the stress response
[@Meeuwisse2007; @Wiese-Bjornstal2009].
For example, a large body of research indicates that training-related stress is also likely to be related to the stress response and injury occurrence [@Lee2017; @Djaoui2017],
and may account for the unexplained variance from the psychological predictors of injury.
Appaneal and Perna [@Appaneal2014]
proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen’s [@Williams1998] model and to address some of these issues.
To elaborate, the BMSAIH aimed to clarify the mediating pathways between the stress response and injury, consider other health outcomes and behavioural factors that impact sports participation,
and integrate the impact of training on athletes’ health [@Appaneal2014].
The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and “widen the window of susceptibility” [@Appaneal2014, p. 74] to a range of undesirable health outcomes including illness and injury.
Consequently, the BMSAIH provides a framework for future research to build on Williams and Andersen's [@Williams1998] model, by including other physiological markers of training-related stress, which together may provide greater insight into the injury process.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence [@Perna1995; @Perna1997; @Perna2003], other research has identified additional markers of training-related stress that are associated with an increased risk of injury; for example,
heart rate variability [@Bellenger2016; @Williams2017],
postural stability [@Romero-Franco2014]
and muscle stiffness [@Pruyn2015].
However, these markers are often studied in isolation without an assessment of the psychosocial factors that are known to contribute to injury, thereby limiting our understanding of how psychosocially and physiologically derived stress may contribute synergistically to injury occurrence.
Recently, Bittencourt et al. [@Bittencourt2016]
suggested that to better understand the multifactorial nature of sports injuries, research needs to move away from studying risk factors in isolation and instead adopt a complex systems approach to injury.
Such an approach posits that injury may arise from a complex “web of determinants” [@Bittencourt2016, p. 3],
where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury.

A challenge when adopting a complex systems approach is using an appropriate analysis technique that is able to capture the uncertainty and complexity of the relationships between different variables. One technique that provides a solution to this challenge is Bayesian network (BN) modelling. BN's allow the construction of graphical probabilistic models using the underlying structure that connects different variables (nodes in the network)  [@Scutari2014]. The learned structure can be used for inference by obtaining the posterior probabilities of a particular node for a given query (e.g., if the value of Node A is x and the value of Node B is y, what is the probability Node C being value z?). Furthermore, BN's do not distinguish between dependent and independent variables as they are a form of unsupervised learning, which is a strength over regression or structure equation models when the underlying relationship in the network may not be known [@Olmedilla2018].

To summarise, despite offering a possible framework to build on the research stemming from Williams and Andersen's [@Williams1998] model, there remains and opportunity to explore other physiological stress-related markers proposed by the BMSAIH, in addition to the already well-established psychological characteristics known to be related to injury [@Appaneal2014].
Furthermore, research has typically not captured changes in both psychosocial factors and stress-related physiological markers that may occur between initial measurement and injury occurrence.
Given the exploratory nature of the current study the following objectives were defined:

- Identify suitable markers of stress that can be easily captured in a large cohort of athletes in a field based setting. 
- Capture the markers of stress and injury occurrence using a prospective, repeated measures design.
- Explore and evaluate the relationships between the markers of stress and injury using Bayesian network modelling.

# Methods {-}

@@

## Participants

A total of 351 athletes (male: _n_ = 231, female: _n_ = 120) were initially recruited for the study. Participants had an average age of `r mean_sd(parchar$age, digits = 1)` years and represented a range of team (football, rugby, netball, cricket, lacrosse, basketball and field hockey) and individual sports (athletics, tennis, weightlifting, gymnastics, judo, swimming and golf from a British University and local sports clubs (Table \ref{tab:table1}).
Participants self-rated competitive level ranged from recreational to international standard. 
A total of `r n_perc(parchar$pic == "1")` participants had sustained an injury in the 12 months prior to the start of the study (male: _n_ = 114 [49%], female; _n_ = 48 [40%]).
At the start of the study, all participants were injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks).
Participants were engaged in training for their respective sports for at least five hours per week.
Ethical approval was obtained from the University ethics committee prior to the start of the study and all participants provided informed consent.

@@

@@

\begin{figure}[!h]
\caption{{\bf mBESS positions (A-F).}
Top row, firm surface. Bottom row, foam surface. Left column, dominant leg stance. Middle column, non-dominant leg stance. Right column, Tandem leg stance.}
\label{fig:fig2}
\end{figure}

The error identification criteria from the original BESS protocol was used by the lead researcher who scored all the BESS trials. 
One error was recorded if any of the following movements were observed during each trial: a) lifting hands off iliac crests; 
b) opening eyes; 
c) stepping, stumbling, or falling; 
d) moving the thigh into more than 30 degrees of flexion or abduction; 
e) lifting the forefoot or heel; and 
f) remaining out of the testing position for more than 5-seconds [@Riemann1999d].
A maximum score of 10 errors was possible for each stance. Multiple errors occurring simultaneously were recorded as one error.
A participant was given the maximum score of 10 if they remained out of the stance position for more than 5-seconds.
To calculate limb asymmetry, the DL and NDL leg score was calculated by summing the DL and NDL errors respectively.
A total score was calculated by summing the total number of errors recorded on all stances (DL, NLD and TS, on foam and firm surfaces). 
To assess the intra-rater reliability, a single measurement, absolute agreement, two-way mixed effects model for the intraclass correlation [ICC; @Koo2016] was used on a sample of 40 participants from the first time point.
The test-retest scoring of BESS resulted in a "good" to "excellent" ICC score (ICC = `r round(icc[["value"]], digits = 2)`, 95% confidence interval = `r paste(round(icc[["lbound"]], digits = 2), "-", round(icc[["ubound"]], digits = 2)) `), indicating the scoring was reliable.

Participants self-reported any injuries they sustained at each data collection during the study period.
An injury was defined as any sports related medical problem causing the athlete to miss or modify their usual training routine. 
Minor scrapes and bruises that may require certain modifications (e.g., strapping or taping) but did not limit continued participation were not considered injuries [cf. @Appaneal2009].
Injury status (did / did not sustain an injury) served as the main outcome measure.

@@

At the start of the UK academic year (September 2016 and 2017), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. 
With the coaches' permission, the lead researcher attended training sessions to inform the athletes about the overall purpose of the study and the requirements of participation.
To be eligible for the study athletes had to be injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) and training a minimum of five hours per week. 
Athletes who met the criteria and volunteered to take part in the study were invited to attend scheduled testing sessions.

A repeated measures prospective cohort design was used to assess athletes' major life events, stress-related physiological markers and injury status over two consecutive twelve-month periods between September 2016 and September 2018. 
Each participant was asked to attend a total of four data collections over a twelve-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). 
Participants provided informed consent before data collection commenced.

\begin{figure}[!h]
\caption{{\bf Study design.}
For each time point (T), each box contains the number of participants who completed the data collection (N), the measures used for data collection and the approximate date of the data collection.}
\label{fig:fig3}
\end{figure}

@@

\begin{figure}[!h]
\caption{{\bf Session protocol.}
Outline of the protocol for each data collection.}
\label{fig:fig4}
\end{figure}

@@

\begin{figure}[H]
\caption{{\bf Example network.}
A simple discrete network contain nodes, possible states of the nodes and the arcs connecting nodes.}
\label{fig:fig5}
\end{figure}

@@

BNs have recently been used in the sport psychology literature [@Olmedilla2018; @Fuster-Parra2017] and offer several benefits over traditional statistical analysis. 
For example, predictions can be made about any variable in the network, rather than there being a distinction between dependent and independent variables in the data, such as in linear regression models that are often used within the sport psychology literature [@Olmedilla2018; @Bittencourt2016].
Furthermore, the structure of a network can be obtained from both empirical data _and_ prior knowledge about the area of study; the latter being particularly useful when there are a large number of variables in the network, or only a small number of observations are available in the data [@Xiao-xuan2007].
In such instances, a purely data driven approach to learning the network would be time-consuming due to the large parameter space, and inefficient at identifying an approximation of the true network structure. 
Prior knowledge about dependencies between variables can therefore be included in the network structure, while still allowing a data driven approach for unknown dependencies, to improve the overall computation of the network structure [@Heckerman1995; @Xu2015].
The following sections detail the steps taken in the current study to firstly prepare the data for the network, and then obtain the structure of the network that was used for inference.

### Data Preparation

To prepare the data for the BN, missing values in the dataset were first imputed. 
Out of the 668 total measurements across all time points in the current study, there were `r n_perc(missingtot$stiffness == 0)` missing muscle stiffness measurements and `r n_perc(missingtot$rmssd == 0)` missing heart rate recordings. The missing data were due to technical faults in the data collection equipment and were considered to be missing completely at random. 
A missing rate of 15-20% has been reported to be common in psychological studies, and several techniques are available to handle missing values [@Enders2003; @Lang2014].
In the current study, the _caret_ package [@Kuhn2008] was used to impute the missing values. 
A bagged tree model using all of the non-missing data was first generated and then used to predict each missing value in the dataset. 
The bagged tree method is a reliable and accurate method for imputing missing values in data and is superior to other commonly used methods such a median imputation [@Kuhn2008].

Preliminary correlation analysis of the data revealed no strong linear dependencies in the data, and therefore the data were binarised to create "Low" and "High" levels within each variable.
Binarising variables is a common procedure for BN modelling and can help improve the fit of the model by reducing the number of possible states for each node [@Beuzen2018].
All variables were approximately normally distributed; therefore, the median of each variable was used as the cut-off point to create the "Low" and "High" levels.

For the LESCA, a cumulative total of the current, and previous time points was calculated at each time point to account for the potential continuing effect of the life events experienced by athletes over time. 
Given the limited support for a relationship between positive life events and injury [@Williams2007], only negative and total life events were included in the network.
Cumulative negative and cumulative total life event scores were first log scaled and then binarised based on the median at each time point (nlelg and tlelg respectively).
In addition to the log scaled cumulative values, an untransformed NLE score from the first time point was included as an additional variable based on previous literature that indicates this variable should have a strong relationship with injury outcome [@Ivarsson2017].

### Network structure

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference. 
Prior knowledge about the network structure was included by providing a list of arcs that are always _restricted_ from being in the network (blacklist), and a list of arcs that are always _included_ in the network (whitelist). 
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded [@Scutari2014]. 
To account for the repeated measures design and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure [@Murphy2002].
In the 2TBN, variables measured T2 could depend on variables measured at T1 (e.g., T1 $\rightarrow$ T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 $\rightarrow$ T3).
However, arcs were blacklisted between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2 to preserve the order in which data was collected. 
Variables were separated into two groups; "explanatory", for variables that did not change during the study (e.g., gender), or "independent", for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with _1 for time point T, and _2 for time point T+1 (e.g., T1_1 $\rightarrow$ T2_2 and T2_1 $\rightarrow$ T3_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis. 
Table \ref{tab:table2} provides an example of the formatted data. 
Participants 1 and 3 have complete data, and therefore have two rows of data each representing variables from T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3, respectively. 
Participant 2 did not complete the final data collection at T3 and therefore only has one row of data representing the variables collected at T1 and T2.
In addition to the blacklisted arcs between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent $\rightarrow$ explanatory); however, arcs were not restricted between explanatory $\rightarrow$ independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables. 
Arcs from clevel $\rightarrow$ gender, nlebase $\rightarrow$ gender and nlebase $\rightarrow$ ind_team were included in the blacklist, as arcs in these directions did not make logical sense. 
All subsequent models used the same blacklist.

```{r table2, results='asis'}
exampledata <- tibble(
  Participant = c("1", "1", "2", "3", "3"),
  "X_1" = c("T1", "T2", "T1", "T1", "T2"),
  " " = "->",
  "X_2" = c("T2", "T3", "T2", "T2", "T3")
)

kable(
  exampledata,
  caption = "Example of the data arrangement used for the network.",
  escape = TRUE
) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0, bold = TRUE)
```

### Preliminary network structures

>>>>

---
title: "wordmaker"
output: bookdown::html_document2
always_allow_html: true

---

@@

```{r include=FALSE}
options(scipen=999, digits = 2)
knitr::opts_chunk$set(echo = FALSE)
# source("r-files/functions.R")
# source("r-files/methods_code.R")
# source("r-files/results_code.R")
load("all_data.RData")
pacman::p_load(tidyverse, knitr, kableExtra, qwraps2, brms, broom.mixed, bnlearn, tidyselect, Rgraphviz)
library(readxl)
library(janitor)
library(irr)
library(lavaan)
library(caret)
library(parallel)
library(zoo)
library(sjmisc)

\section{Introduction}

Over the last four decades sport-related injuries have received increased research attention [@Ivarsson2017] in response to the high incidence [@Rosa2014; @Sheu2016] and associated undesirable physical and psychological effects [@Leddy1994; @Brewer2012]. Multiple psychological [@Slimani2018], anatomical [@Murphy2003], biomechanical [@Neely1998; @Hughes2014] and environmental [@Meeuwisse2007] factors have been associated with sports injury occurrence and several models of injury causation have been proposed that highlight the multifactorial nature of injury occurrence [@Kumar2001; @Meeuwisse2007; @Wiese-Bjornstal2009], of which one of the most widely cited was developed by Williams and Anderson [Fig \ref{fig:fig1}; @Andersen1988; @Williams1998].

Williams and Andersen’s [@Williams1998] stress-injury model proposed that when faced with a potentially stressful athletic situation, an athlete’s personality traits (e.g., hardiness, locus of control and competitive trait anxiety), history of stressors (e.g., major life events and previous injuries) and coping resources (e.g., general coping behaviours) contribute to the injury response, either interactively or in isolation. The stress response is central to the model and reflects the bi-directional relationship between athletes’ appraisal of, and response to, a stressful athletic situation. The model predicts that athletes who have a history of stressors, personality traits that intensify the stress response and few coping resources, will exhibit greater attentional (e.g., peripheral narrowing) and/or physiological (e.g., increased muscle tension) responses that put these individuals at greater injury risk.

Within Williams and Andersen’s [@Williams1998] model, major life events, which is a component of an athlete’s history of stressors, most consistently predicts injury occurrence [@Williams2007]; specifically, major life events with a negative, as opposed to positive, valence [@Passer1983a; @Maddison2005]. Personality traits and coping resources have also been found to predict injury with athletes more likely to sustain an injury if they have poor social support and psychological coping skills, and high trait anxiety and elevated competitive state anxiety; compared to those with the opposing profile [@Smith1990; @Lavallee1996; @Ivarsson2010].However, the amount of variance explained by these psychosocial factors has been modest and typically between 5 - 30% [@Galambos2005; @Ivarsson2010], which indicates a likely interaction with other factors.

While the psychosocial factors in Williams and Andersen’s [@Williams1998] model have received the most research attention, less insight into the mechanisms through which these factors are proposed to exert their effect exists. To elaborate, the model suggests that injuries are likely to occur through either increased physiological arousal resulting in increased muscle tension and reduced flexibility or attentional deficits caused by increased distractibility and peripheral narrowing. However, to date, the research has largely focused on attentional deficits [@Andersen1999; @Rogers2005; @Wilkerson2012a; @Swanik2007]. For example, Andersen and Williams [@Andersen1999] found athletes with high life event stress coupled with low social support had greater peripheral narrowing under stressful conditions compared to athletes with the opposing profile; these athletes went on to sustain an increased number of injuries during the following athletic season. Rodgers and Landers [@Rogers2005] further supported Andersen and Williams’s [@Andersen1999] earlier findings by identifying that peripheral narrowing under stress mediated 8.1% of the relationship between negative life events and injury.

Knowledge of the physiological factors (e.g., increased muscle tension and reduced motor control) contributing to the remaining variance between negative life events and athletic injury remains sparse [cf. @Williams1998]. One challenge faced by researchers addressing the sports injury problem through a psychological lens is the multifactorial nature of injury, and the possible interaction with physiological factors in the stress response [@Meeuwisse2007; @Wiese-Bjornstal2009]. For example, a large body of research has suggested that training-related stress is also likely to contribute to the stress response and injury occurrence [@Lee2017; @Djaoui2017] and may account for the unexplained variance from the psychological predictors. Appaneal and Perna [@Appaneal2014] proposed the biopsychosocial model of stress athletic injury and health (BMSAIH) to serve as an extension to Williams and Andersen’s [@Williams1998] model. The BMSAIH enhances our understanding of the mediating pathways between the stress response and injury alongside other health outcomes and behavioral factors that impact sports participation [@Appaneal2014]. The central tenet of the BMSAIH is that psychosocial distress (e.g., negative life events) may act synergistically with training-related stress as a result of high-intensity and high-volume sports training, and “widen the window of susceptibility” [@Appaneal2014, p. 74] to a range of undesirable health outcomes including illness and injury. Consequently, the BMSAIH provides an important framework that has enhanced insight into the multi-faceted nature of the injury process by building on Williams and Andersen’s [@Williams1998] model whilst including other physiological markers of training-related stress.

Although research supporting the BMSAIH has mainly focused on the relationship between hormonal responses to training and injury occurrence [@Perna1995; @Perna1997; @Perna2003], other training-related stress markers including heart rate variability [@Bellenger2016; @Williams2017],  muscle stiffness [@Pruyn2015] and postural stability [@Romero-Franco2014] have been reported to be associated with an increased risk of injury. However, the largely mono-disciplinary approach to examining the role of these markers has provided limited insight into the multifaceted interactions with psychological factors that may contribute to injury occurrence. Furthermore, a reliance on designs that capture a single point of measurement precludes the assessment of intra as well as inter-individual changes and the effect of the time interval between measurement and injury occurrence on subsequent injuries [@Johnson2014]. Such an approach fails to capture changes in both psychological factors and physiological markers that may occur preceding an injury. Importantly, a repeated measurement approach would enable an assessing of how variables and their interactions change with respect to time provide greater insight into the effect that repeated exposure to major life events and other stress-related factors has on injury occurrence.

Recently, Bittencourt et al. [@Bittencourt2016] advocated a move away from studying isolated risk factors and instead, adopt a complex systems approach in order to understand injury occurrence. Such an approach posits that injury may arise from a complex “web of determinants” [@Bittencourt2016, p. 3], where different factors interact in unpredictable and unplanned ways, but result in a global outcome pattern of either adaptation or injury. Capturing the uncertainty and complexity of the relationships between different variables using an appropriate interdisciplinary analysis within the framework of a complex systems approach is challenging. Bayesian network (BN) modelling provides one solution by allowing the construction of graphical probabilistic models using the underlying structure that connects different variables [@Scutari2014]. The learned BN structure can be used for inference by obtaining the posterior probabilities of a particular variable for a given query (e.g., if the value of variable A is x and the value of variable B is y, what is the probability variable C of being value z?). Furthermore, unlike regression or structural equation models, BN’s do not distinguish between dependent and independent variables when the underlying relationship in the network may not be known [@Olmedilla2018]. BN modelling subsequently provides a valuable but underused interdisciplinary approach to investigating the complex and unpredictable interactions of psychological and physiological factors implicated in the injury process.

Using the frameworks provided by Williams and Andersen’s [@Williams1998] stress injury model and Appaneal’s BMSAIH model, the aim of this interdisciplinary study was to develop new understanding of the multifaceted interactions of psychological and physiological stress-related factors with injury occurrence. A prospective, repeated measures design incorporating field-based physiological and psychological measures combined with a BN modelling analysis was used to address the study aim.

\section{Methods}

@@

\subsection{Participants}

A total of 351 athletes (male: _n_ = 231, female: _n_ = 120) were initially recruited for the study from a British University and local sports clubs (Table \ref{tab:table1}). Participants had an average age of `r mean_sd(parchar$age, digits = 1)` years and represented a range of team (football, rugby, netball, cricket, lacrosse, basketball and field hockey) and individual sports (athletics, tennis, weightlifting, gymnastics, judo, swimming and golf. 
Participants self-rated competitive level ranged from recreational to international standard. 
They were engaged in training for their respective sports for at least five hours per week.
A total of `r n_perc(parchar$pic == "1")` participants had sustained an injury in the 12 months prior to the start of the study (male: _n_ = 114 [49%], female; _n_ = 48 [40%]).
All participants were injury free (no modifications to their usual training routine due to a sport related medical problem for a minimum of four weeks) at the start of the study. 
Ethical approval was obtained from the University ethics committee prior to the start of the study and all participants provided informed consent.

@@

Participants self-reported any injuries they sustained at each data collection during the study period.
An injury was defined as any sports related medical problem causing the athlete to miss or modify their usual training routine for at least 24 hours [@Fuller2006; @Fuller2007b; @Timpka2014]. 
Minor scrapes and bruises that may require certain modifications (e.g., strapping or taping) but did not limit continued participation were not considered injuries [cf. @Appaneal2009].
Injury status (did / did not sustain an injury) served as the main outcome measure.

@@

The error identification criteria from the original BESS protocol was used by the lead researcher who scored all the BESS trials. 
One error was recorded if any of the following movements were observed during each trial: a) lifting hands off iliac crests; 
b) opening eyes; 
c) stepping, stumbling, or falling; 
d) moving the thigh into more than 30 degrees of flexion or abduction; 
e) lifting the forefoot or heel; and 
f) remaining out of the testing position for more than 5-seconds [@Riemann1999d].
A maximum score of 10 errors was possible for each stance. Multiple errors occurring simultaneously were recorded as one error.
A participant was given the maximum score of 10 if they remained out of the stance position for more than 5-seconds.
A total score was calculated by summing the total number of errors recorded on all stances (DL, NLD and TS, on foam and firm surfaces). 
To assess the intra-rater reliability, a single measurement, absolute agreement, two-way mixed effects model for the intraclass correlation [ICC; @Koo2016] was used on a sample of 40 participants from the first time point.
The test-retest scoring of BESS resulted in a "good" to "excellent" ICC score (ICC = `r round(icc[["value"]], digits = 2)`, 95% confidence interval = `r paste(round(icc[["lbound"]], digits = 2), "-", round(icc[["ubound"]], digits = 2)) `), indicating the scoring was reliable.

@@

At the start of the academic year (September), coaches of sports teams at a British University and local sports clubs were contacted and informed about the study. With the coaches’ permission, the lead researcher attended training sessions to inform athletes about the overall purpose of the study and the requirements of participation. Athletes who met the participation criteria and volunteered to take part in the study were invited to attend scheduled testing sessions. A repeated measures prospective cohort design was used to assess athletes’ major life events, stress-related physiological markers and injury status over two consecutive 12-month periods. Each participant was asked to attend a total of three data collections over a 12-month period, with each data collection separated by a four-month interval (Fig \ref{fig:fig3}). Participants provided informed consent before data collection commenced.

@@

@@

In order to use discrete networks, continuous variables must first be split into categorical levels.
When there are a large number of variables in the network, limiting the number of levels has the benefit of producing a network that is more parsimonious in terms of parameters. For example, a network with 10 variables each with two levels has 100 (10^2) possible parameter combinations, however the same network with three levels has 1000 (10^3) possible  parameter combinations, the latter being significantly more computationally expensive. Using a larger number of splits in the data also comes at a cost of reducing the statistical power in detecting probabilistic associations, and reduces the precision of parameter estimates for the probabilistic associations that are detected because it reduces the sample-size-to-parameters ratio [@Scutari2014]. Typically, no more than three levels have been used in Bayesian networks in the sports injury literature [@Olmedilla2018; @Ruiz-Perez2019]

@@

BNs have recently been used in the sport psychology literature [@Olmedilla2018; @Fuster-Parra2017] and offer several benefits over traditional statistical analysis. 
For example, predictions can be made about any variable in the network, rather than there being a distinction between dependent and independent variables in the data, such as in linear regression models that are often used within the sport psychology literature [@Olmedilla2018; @Bittencourt2016].
Furthermore, the structure of a network can be obtained from both empirical data _and_ prior knowledge about the area of study; the latter being particularly useful when there are a large number of variables in the network, or only a small number of observations are available in the data [@Xiao-xuan2007].
In such instances, a purely data driven approach to learning the network would be time-consuming due to the large parameter space, and inefficiency at identifying an approximation of the true network structure. 
Prior knowledge about dependencies between variables can therefore be included in the network structure, while still allowing a data driven approach for unknown dependencies, to improve the overall computation of the network structure [@Heckerman1995; @Xu2015].
The following sections detail the steps taken in the current study to firstly prepare the data for each network, and then obtain the structure of each network that was used for further inference.

### First network

#### Data Preparation

Of the 351 participants that were initially recruited for the study, 94 only completed the first time point, and therefore had to be removed from the study as no injury information was obtained for these participants following the first time point. 
To prepare the data for the BN, missing values in the dataset were first imputed. 
Out of the 650 total measurements across all time points in the current study, there were `r n_perc(missingtot$stiffness == 0)` missing muscle stiffness measurements and `r n_perc(missingtot$rmssd == 0)` missing heart rate recordings. The missing data were due to technical faults in the data collection equipment and were considered to be missing completely at random. 
A missing rate of 15-20% has been reported to be common in psychological studies, and several techniques are available to handle missing values [@Enders2003; @Lang2014].
In the current study, the _caret_ package [@Kuhn2008] was used to impute the missing values. 
A bagged tree model using all of the non-missing data was first generated and then used to predict each missing value in the dataset. 
The bagged tree method is a reliable and accurate method for imputing missing values in data and is superior to other commonly used methods such a median imputation [@Kuhn2008].

A median split technique was used to discretise the data used in the network into "Low" and "High" levels.
All variables apart from negative and total life events were approximately normally distributed (based on visual inspection, see supplementary material Sx) and required no further transformation prior to the median split. 
For the LESCA questionnaire data, a cumulative total of the current, and previous time points was calculated at each time point to account for the potential continuing effect of the life events experienced by athletes over time.
Given the limited support for a relationship between positive life events and injury [@Williams2007], only negative and total life events were included in the network.
Cumulative negative, and cumulative total life event scores at each time point were first log scaled so distributions were approximately normal, and then binarised using the median at each time point (nlelg and tlelg respectively).
In addition to the log scaled cumulative values, an untransformed NLE score from the first time point was included as an additional variable based on previous literature that indicates this variable should have a strong relationship with injury outcome [@Ivarsson2017].

#### Network structure

To obtain the network structure, several steps were taken to ensure that both a theoretically realistic network, and a network that was an appropriate fit to the collected data, was used for inference. 
Prior knowledge about the network structure was included by providing a list of arcs that are always _restricted_ from being in the network (blacklist), and a list of arcs that are always _included_ in the network (whitelist). 
Additionally, there are several scoring functions such as Bayesian Information Criteria (BIC) and Bayesian Dirichlet equivalent uniform (BDeu) that can be used to compare network structures with certain nodes and arcs included or excluded [@Scutari2014]. 
To account for the repeated measures design employed in this study and to maximise the use of the data, pairs of complete cases (e.g., participants who completed T1 + T2, and T2 + T3) were used in a two-time Bayesian network (2TBN) structure [@Murphy2002].
In the 2TBN, variables measured T2 could depend on variables measured at T1 (e.g., T1 $\rightarrow$ T2) and variables measured at T3 could depend on variables measured at T2 (e.g., T2 $\rightarrow$ T3).
However, arcs were blacklisted between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2 to preserve the order in which data was collected. 
Variables were separated into two groups; "explanatory", for variables that did not change during the study (e.g., gender), or "independent", for variables that were measured at each time point and could vary during the study.
Independent variable names were suffixed with _1 for time point T, and _2 for time point T+1 (e.g., T1_1 $\rightarrow$ T2_2 and T2_1 $\rightarrow$ T3_2).
Formatting the data in this way meant participants who completed T1 and T2, but did not complete T3, could still be included in the analysis. 
Table \ref{tab:table2} provides an example of the formatted data and demonstrates that participants 1 and 3 have complete data, and therefore have two rows of data each representing variables from T1 $\rightarrow$ T2 and T2 $\rightarrow$ T3, respectively. 
Participant 2 did not complete the final data collection at T3 and therefore only has one row of data representing the variables collected at T1 and T2.
In addition to the blacklisted arcs between T2 $\rightarrow$ T1 and T3 $\rightarrow$ T2, the direction of arcs was restricted between independent variables and explanatory variables (e.g., independent $\rightarrow$ explanatory); however, arcs were not restricted between explanatory $\rightarrow$ independent variables.
Finally, arc direction was restricted between specific nodes within the explanatory variables. 
Arcs from clevel $\rightarrow$ gender, nlebase $\rightarrow$ gender and nlebase $\rightarrow$ ind_team were included in the blacklist, as arcs in these directions did not make logical sense. 
All subsequent models used the same blacklist.

```{r table2, results='asis'}
exampledata <- tibble(
  Participant = c("1", "1", "2", "3", "3"),
  "X\\_1" = c("T1", "T2", "T1", "T1", "T2"),
  "X\\_2" = c("T2", "T3", "T2", "T2", "T3")
)

kable(
  exampledata,
  caption = "Example of the data arrangement used for the network.",
  escape = FALSE
) %>%
  kable_styling(latex_options = "HOLD_position") %>%
  row_spec(0, bold = TRUE)
```

#### Preliminary network structures

Prior to the final network structure presented in the results section, several structures with different combinations of variables were explored. Networks were learned using a Tabu search algorithm [@Norvig2009] and BIC was used to compare different models. A higher BIC value indicates the structure of a DAG is a better fit to the observed data [@Scutari2014]. BIC values for each combination of variables of interest are reported as the combination of variables with the highest BIC value, followed by the relative scores of the other variables in the model.
